---
title: "Group Project - Flight Demand"
author: "Femi Fabiyi, Olivia Fenwick, Todd Livergood"
date: "Spring Quarter - 2021"
output: html_document
---

### Initiate data and libraries

```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

#### Load libraries

```{r}
library("tseries")
library(corrplot)
source("http://www.sthda.com/upload/rquery_cormat.r")
library(olsrr)
library(Metrics)
library(imputeTS)
library(forecast)
library(prophet)
library(bsts)
library(dplyr)
library(plotly)
```

#### Read in data:

```{r}
airport_pull <- read.csv("Airport Final.csv")
airport_pull_df <- as.data.frame(airport_pull)
head(airport_pull_df)

airport_data_original <- read.csv("Airport Data.csv", header=TRUE)
#head(airport_data_original)
```

#### Eliminate routes that haven't flown in past year and impute zeros:

```{r}
# remove routes that haven't flown in past year
airport_df <- subset(airport_pull_df, select=c('YYYYMM'))
colname <- c('YYYYMM')
airport.routes <- colnames(airport_pull_df)
for (i in 2:(length(airport_pull_df)-1)){
  if (sum(tail(airport_pull_df[,i], 12))!=0){
    colname <- c(colname, airport.routes[i])
    airport_df <- cbind(airport_df, airport_pull_df[,i])
  }
}
colnames(airport_df) <- colname

#head(airport_df)
```

#### Impute zero to value we would have expected demand to be if the flight had flown

```{r}
# fill 0 with NA
airport_df[airport_df==0] <- NA

# fill missing values through linear interpolation
airport_df <- na_interpolation(airport_df, option='linear')
#head(airport_df)
```

#### Split the data into a training and test set (train: 2000-2018, test: 2019)

```{r}
# train test split
train <- head(airport_df, n = 240 - 12)
test <- tail(airport_df, n=12)

# convert to time series
train_ts <- ts(train, frequency = 12)
test_ts <- ts(test, frequency = 12)
```

### Exploratory data analysis

#### Graph data to see cutoff:
```{r}
# view all routes
matplot(airport_df, type="l")

# view biggest route
matplot(airport_df$MELBOURNE.SYDNEY, type="l")

# view last 24 months (2018/2019) of biggest route
airport_df_tail <- tail(airport_df,24)
matplot(airport_df_tail["MELBOURNE.SYDNEY"], type="l")
```

#### Graph total passengers count by year

```{r}
# extract total passengers for each year
all_total_passengers_per_year <- airport_data_original %>%
  dplyr::select(City.Pair, Year, Passenger.Total)%>%
  arrange(City.Pair) %>%
  group_by(Year) %>% summarize(sum(Passenger.Total))#!!sym(alias)

# graph data
fig <- plot_ly(all_total_passengers_per_year, x = ~Year, y= ~`sum(Passenger.Total)`, color = ~Year, type =  "bar", alpha=0.8)
fig
```

#### Graph total passengers count by year by route

```{r}
# extract total passengers across all routes for specific year
total_passenger_by_route_for_year <- airport_data_original %>% arrange(Year)%>% 
  dplyr::select(City.Pair, Year, Passenger.Total)%>% 
  group_by(City.Pair, Year)%>% 
  summarize(total_route_flights = sum(Passenger.Total))

# graph data
fig <- plot_ly(total_passenger_by_route_for_year, x = ~Year, y = ~total_route_flights, color = ~total_route_flights, type =  "box", alpha=0.8)
fig
```

#### Graph total passenger by route by month for 2019

```{r}
# extract total passengers across all routes for specific year
total_passenger_by_route_for_specific_year <- airport_data_original %>% arrange(Year)%>% 
  dplyr::select(City.Pair, Year, Passenger.Total)%>% 
  group_by(City.Pair, Year)%>% 
  summarize(total_route_flights = sum(Passenger.Total))%>%
  filter(Year==2019)

# graph data
fig <- plot_ly(total_passenger_by_route_for_year, x = ~Year, y = ~total_route_flights, color = ~City.Pair, type =  "box", alpha=0.8)
fig
```

#### Graph total pax by route by year with error bars

```{r}
# extract total number of passengers for specific route across for each year
yearly_total_passenger_by_route <- airport_data_original %>%
  dplyr::select(City.Pair, Year, Passenger.Total)%>%
  arrange(City.Pair) #%>% filter( City.Pair == "ADELAIDE-ALICE SPRINGS")

# graph data
fig <- plot_ly(yearly_total_passenger_by_route,x=~Year, y = ~Passenger.Total, color = ~City.Pair, type =  "box", alpha=0.8)
fig
```

#### Graph total passengers by month and year with route filter

```{r}
# extract total number of passengers for a specific route across the period for each month
monthly_total_passenger_specific_route <- airport_data_original %>%
  dplyr::select(City.Pair, Year, Month_num, Passenger.Total)%>%
  arrange(City.Pair) #%>% filter( City.Pair == "ADELAIDE-ALICE SPRINGS")

# graph data
fig <- plot_ly(monthly_total_passenger_specific_route,x=~Year, y = ~Month_num, text = ~Passenger.Total, color = ~City.Pair, type =  "scatter")
fig
```

#### Run adf and kpss tests on each of the series:

```{r}
# initiate test summary
test_summary <- matrix(ncol=4, nrow=length(airport_df))
for (i in 2:length(airport_df)) {
  aaa <- adf.test(airport_df[,i])
  aaa_conc <- if (aaa$p.value <= .05) {"stationary"} else {"non-stationary"}
  bbb <- kpss.test(airport_df[,i])
  bbb_conc <- if (bbb$p.value <= .05) {"non-stationary"} else {"stationary"}
  # update test summary
  test_summary[i-1,] <- c(aaa$p.value,aaa_conc,bbb$p.value,bbb_conc)
}
test_summary <- as.data.frame(test_summary)
colnames(test_summary) <- c("adf p-value","adf result","kpss p-value","kpss result")
 
head(test_summary)
```

#### Summarize freq of test values (adf test):

```{r}
adf_summary <- table(test_summary$`adf result`)
adf_summary = as.data.frame(adf_summary)

adf_summary
```


#### Summarize freq of test values (kpss test):

```{r}
kpss_summary <- table(test_summary$`kpss result`)
kpss_summary = as.data.frame(kpss_summary)

kpss_summary
```

#### View acf of most pupular route:

```{r}
acf(airport_df["MELBOURNE.SYDNEY"])
```

#### Correlation matrix (Note: Commenting this out as it was printing multiple pages when knitting.)
```{r}
# airport_df_temp = subset(airport_df, select = -c(YYYYMM))
# rquery.cormat(airport_df_temp)
```



### Build models

```{r}
routes <- colnames(train_ts)
routes <- tail(routes, length(routes)-1)

# initiate model summary table
model_aa_summary <- data.frame(route = character(), model_name = character(), AIC = double(), AICc = double(), BIC = double(), sMAPE = double())
```

#### auto.arima 

```{r}
# create future forecast dataframe
SARIMA.forecast <- data.frame(c(1,2,3,4,5,6,7,8,9,10,11,12,1,2,3,4,5,6,7,8,9,10,11,12))
names(SARIMA.forecast) <- c("Month")
counter <- 2

for (i in routes) {
  # calculate auto.arima for route i
  route_x <- train_ts[,i]
  model_aa_x <- auto.arima(route_x)
  res_x <- model_aa_x$residuals
  res_x_ts <- ts(res_x)
  route <- i
  # save AIC/AICc/BIC
  aic <- model_aa_x$aic
  aicc <- model_aa_x$aicc
  bic <- model_aa_x$bic
  # generate model forecast
  fc_x <- as.data.frame(forecast(model_aa_x),h=12)
  model_aa_forecast <- fc_x$`Point Forecast`
  # update forecast dataframe
  forecast.i.df <- as.data.frame(forecast(model_aa_x),h=24)
  forecast.i <- forecast.i.df$`Point Forecast`
  SARIMA.forecast <- cbind(SARIMA.forecast, forecast.i)
  colnames(SARIMA.forecast)[counter] <- route
  counter = counter + 1
  # calculate sMAPE
  sMAPE_x <- smape(test[,i],model_aa_forecast)
  # update model summary tale
  model_aa_summary <- rbind(model_aa_summary, c(route,"SARIMA",round(aic, digits = 2),round(aicc,digits = 2),round(bic,digits = 2),sMAPE_x))
}

# rename columns of model summary table
colnames(model_aa_summary) <- c("Route", "Model","AIC","AICC","BIC","sMAPE")
head(model_aa_summary)

# update arima forecast to just 2022 forecast
SARIMA.forecast <- tail(SARIMA.forecast, 12)
#head(SARIMA.forecast)
```

#### Holt Winters (multiplicative seasonality and additive seasonality)

```{r}
# create future forecast dataframe
HW.add.forecast <- data.frame(c(1,2,3,4,5,6,7,8,9,10,11,12,1,2,3,4,5,6,7,8,9,10,11,12))
names(HW.add.forecast) <- c("Month")
HW.mult.forecast <- data.frame(c(1,2,3,4,5,6,7,8,9,10,11,12,1,2,3,4,5,6,7,8,9,10,11,12))
names(HW.mult.forecast) <- c("Month")
counter <- 2

for (i in routes) {
  # calculate holt-winters for route i
  route_x <- train_ts[,i]
  model_hw_add <- hw(route_x,seasonal="additive")
  model_hw_mult <- hw(route_x,seasonal="multiplicative")
  route <- i
  # save AIC/AICc/BIC
  aic_add <- model_hw_add$model$aic
  aicc_add <- model_hw_add$model$aicc
  bic_add <- model_hw_add$model$bic
  aic_mult <- model_hw_mult$model$aic
  aicc_mult <- model_hw_mult$model$aicc
  bic_mult <- model_hw_mult$model$bic
  # generate the forecast
  fc_x_add <- as.data.frame(forecast(model_hw_add),h=12)
  model_aa_forecast_add <- fc_x_add$`Point Forecast`
  fc_x_mult <- as.data.frame(forecast(model_hw_mult),h=12)
  model_aa_forecast_mult <- fc_x_mult$`Point Forecast`
  # update forecast dataframe
  forecast.i.df <- as.data.frame(forecast(model_hw_add),h=24)
  forecast.i <- forecast.i.df$`Point Forecast`
  HW.add.forecast <- cbind(HW.add.forecast, forecast.i)
  colnames(HW.add.forecast)[counter] <- route
  forecast.i.df <- as.data.frame(forecast(model_hw_mult),h=24)
  forecast.i <- forecast.i.df$`Point Forecast`
  HW.mult.forecast <- cbind(HW.mult.forecast, forecast.i)
  colnames(HW.mult.forecast)[counter] <- route
  counter = counter + 1
  # calculate sMAPE
  sMAPE_x_add <- smape(test[,i],model_aa_forecast_add)
  sMAPE_x_mult <- smape(test[,i],model_aa_forecast_mult)
  # update the model summary
  model_aa_summary <- rbind(model_aa_summary, c(route, "HW-add.", aic_add, aicc_add, bic_add, sMAPE_x_add))
  model_aa_summary <- rbind(model_aa_summary, c(route, "HW-mult.", aic_mult, aicc_mult, bic_mult, sMAPE_x_mult))
}

tail(model_aa_summary)

# update HW forecast to just 2022 forecast
HW.add.forecast <- tail(HW.add.forecast, 12)
#head(HW.add.forecast)
HW.mult.forecast <- tail(HW.mult.forecast, 12)
#head(HW.mult.forecast)
```

#### Prophet

```{r}
# create future forecast dataframe
prophet.forecast <- data.frame(c(1,2,3,4,5,6,7,8,9,10,11,12))
names(prophet.forecast) <- c("Month")
counter <- 3

for (i in routes) {
  # calculate prophet for route i
  model_prophet_x <- data.frame(ds = seq(as.Date('2000-01-01'),as.Date('2018-12-01'), by = 'm'), y = train[,i])
  m <- prophet(model_prophet_x)
  route <- i
  future <- make_future_dataframe(m, periods = 12, freq = 'month')
  # generate forecast
  forecast <- predict(m, future)
  forecast_tail <- tail(forecast,12)
  # update forecast dataframe
  future2 <- make_future_dataframe(m, periods = 24, freq = 'month')
  forecast2 <- predict(m, future2)
  forecast_tail2 <- tail(forecast2,12)
  prophet.forecast <- cbind(prophet.forecast, forecast_tail2)
  colnames(prophet.forecast)[counter] <- route
  counter = counter + 1
  # calculate sMAPE
  sMAPE_prophet <- smape(test[,i],forecast_tail$yhat)
  # update model summary
  model_aa_summary <- rbind(model_aa_summary, c(route, "Phophet", 999999, 999999, 999999, sMAPE_prophet))
}  

tail(model_aa_summary)

#head(prophet.forecast)
```

### Seasonal Decomposition of Time Series by Loess (STL - only additive and requires box cos to handle multiplicative)

```{r}
# create future forecast dataframe
STL.forecast <- data.frame(c(1,2,3,4,5,6,7,8,9,10,11,12,1,2,3,4,5,6,7,8,9,10,11,12))
names(STL.forecast) <- c("Month")
counter <- 2

for (i in routes) {
  # calculate STL for route i
  s.7 <-stl(train_ts[,i], s.window= 7, robust = TRUE)
  route <- i
  # generate forecast
  forecast_stl <- forecast(s.7, h = 12)
  fc_stl1 <- as.data.frame(forecast_stl)
  model_aa_forecast_stl1 <- fc_stl1$`Point Forecast`
  # update forecast dataframe
  forecast.i.df <- as.data.frame(forecast(s.7, h = 24))
  forecast.i <- forecast.i.df$`Point Forecast`
  STL.forecast <- cbind(STL.forecast, forecast.i)
  colnames(STL.forecast)[counter] <- route
  counter = counter + 1
  # calculate # sMAPE
  sMAPE_x_stl1 <- smape(test[,i],model_aa_forecast_stl1)
  # update model summary
  model_aa_summary <- rbind(model_aa_summary, c(route, "STL", 999999, 999999, 999999, sMAPE_x_stl1))
}  

tail(model_aa_summary)

# update STL forecast to just 2022 forecast
STL.forecast <- tail(STL.forecast, 12)
#head(STL.forecast)
```

#### BSTS

```{r}
# create future forecast dataframe
BSTS.forecast <- data.frame(c(1,2,3,4,5,6,7,8,9,10,11,12,1,2,3,4,5,6,7,8,9,10,11,12))
names(BSTS.forecast) <- c("Month")
counter <- 2

for (i in routes) {
  # calculate BSTS for route i
  route <- i
  y <-log10(train_ts[,i])
  y_test_log <- log10(test[,i])
  ss <- AddLocalLinearTrend(list(), y)
  ss <- AddSeasonal(ss, y, nseasons= 12)
  model_bsts <- bsts(y, state.specification= ss, niter = 100)
  # generate forecast
  pred <- predict.bsts(model_bsts, horizon = 12, quantiles = c(.025, .975))
  # update forecast dataframe
  pred2 <- predict.bsts(model_bsts, horizon = 24, quantiles = c(.025, .975))
  e_pred <- 10^(pred2$mean)
  BSTS.forecast <- cbind(BSTS.forecast, e_pred)
  colnames(BSTS.forecast)[counter] <- route
  counter = counter + 1
  # calculate sMAPE after "untransforming" the data:
  y_test_log_back <- 10^(y_test_log)
  pred_back <- 10^(pred$mean)
  sMAPE_bsts <- smape(y_test_log_back,pred_back)
  
  # update model summary
  model_aa_summary <- rbind(model_aa_summary, c(route, "BSTS", 999999, 999999, 999999, sMAPE_bsts))
}  

tail(model_aa_summary)

# update BSTS forecast to just 2022 forecast
BSTS.forecast <- tail(BSTS.forecast, 12)
#head(BSTS.forecast)
```

### Model comparison

```{r}

model_aa_summary_final <- transform(model_aa_summary, sMAPE = as.numeric(sMAPE))

# mean:
mean_sMAPE <- model_aa_summary_final %>% group_by(Model) %>% summarise_at(c("sMAPE"),mean)
mean_sMAPE$sMAPE <- round(mean_sMAPE$sMAPE ,4)
mean_sMAPE <- mean_sMAPE[order(mean_sMAPE$sMAPE),]

# median:
median_sMAPE <- model_aa_summary_final %>% group_by(Model) %>% summarise_at(c("sMAPE"),median)
median_sMAPE$sMAPE <- round(median_sMAPE$sMAPE ,4)
median_sMAPE <- median_sMAPE[order(median_sMAPE$sMAPE),]
```

#### View graph of models by average

```{r}
# mean
par(mar=c(8,6,6,6))
xx <- barplot(mean_sMAPE$sMAPE, main="Mean sMAPE", xlab="Model", names.arg = mean_sMAPE$Model, las = 2, ylim = c(0,.1))
text(x = xx, y = mean_sMAPE$sMAPE, label = mean_sMAPE$sMAPE, pos = 3, cex = 0.8, col = "red")

# median
par(mar=c(8,6,6,6))
yy <- barplot(median_sMAPE$sMAPE, main="Median sMAPE", xlab="Model", names.arg = median_sMAPE$Model, las = 2, ylim = c(0,.1))
text(x = yy, y = median_sMAPE$sMAPE, label = median_sMAPE$sMAPE, pos = 3, cex = 0.8, col = "red")

```

#### Save forecasts to csv

```{r}
# ARIMA
write.csv(SARIMA.forecast, 'SARIMA forecast.csv', row.names=FALSE)

# HW add
write.csv(HW.add.forecast, 'HW add forecast.csv', row.names=FALSE)

# HW mult
write.csv(HW.mult.forecast, 'HW mult forecast.csv', row.names=FALSE)

# prophet
write.csv(prophet.forecast, 'prophet forecast.csv', row.names=FALSE)

# STL
write.csv(STL.forecast, 'STL forecast.csv', row.names=FALSE)

# BSTS
write.csv(BSTS.forecast, 'BSTS forecast.csv', row.names=FALSE)
```

#### STL Route-specific plots (enter route into ii value):

```{r}
# specify route
ii <- "MELBOURNE.SYDNEY"

s.7 <-stl(train_ts[,ii], s.window= 7, robust = TRUE)
route <- ii
# generate forecast
forecast_stl <- forecast(s.7, h = 12)
fc_stl1 <- as.data.frame(forecast_stl)
model_aa_forecast_stl1 <- fc_stl1$`Point Forecast`
#plot(model_aa_forecast_stl1, main=ii)

# plot actual vs forecast
plot(test[,ii], main=ii, xlab = "Month", ylab = "Passenger Count", type='l')  
lines(model_aa_forecast_stl1, col='red')

```


#### Plot forecasts vs actual test for different models (to include in slides):
```{r}
#i <- "MELBOURNE-SYDNEY"
i <- "MELBOURNE.SYDNEY"
ii <- test$`MELBOURNE.SYDNEY`
#train_ts[,"MELBOURNE-SYDNEY"]

#train_ts

##SARIMA (red):
route_x_chart <- train_ts[,i]
model_aa_x_chart <- auto.arima(route_x_chart)
fc_x_chart <- as.data.frame(forecast(model_aa_x_chart, h=12))

##HW :
model_hw_add_chart <- hw(route_x_chart,seasonal="additive")
model_hw_mult_chart <- hw(route_x_chart,seasonal="multiplicative")

fc_x_chart_hwadd <- as.data.frame(forecast(model_hw_add_chart, h=12))
fc_x_chart_hwmult <- as.data.frame(forecast(model_hw_mult_chart, h=12))

##STL:
s.7_chart <-stl(route_x_chart, s.window= 7, robust = TRUE)
forecast_stl_chart <- forecast(s.7_chart, h = 12)
fc_stl1_chart <- as.data.frame(forecast_stl_chart)


##Prophet:
model_prophet_x_chart <- data.frame(ds = seq(as.Date('2000-01-01'),as.Date('2018-12-01'), by = 'm'), y = train[,i])
m_chart <- prophet(model_prophet_x_chart)
future_chart <- make_future_dataframe(m_chart, periods = 12, freq = 'month')
forecast_prophet_chart <- predict(m_chart, future_chart)
forecast_tail_chart <- tail(forecast_prophet_chart,12)


##BSTS:
y_chart <-log10(route_x_chart)
y_test_log_chart <- log10(test[,i])
ss_chart <-AddLocalLinearTrend(list(), y_chart)
ss_chart <-AddSeasonal(ss_chart, y_chart, nseasons= 12)
model_bsts_chart <-bsts(y_chart, state.specification= ss_chart, niter = 100)
pred_chart<-predict.bsts(model_bsts_chart, horizon = 12, quantiles = c(.025, .975))
e_pred_chart <- 10^(pred_chart$mean)


##Plot the charts:
plot(ii, main=i, ylab = "Passenger Count", xlab = "Month in 2019", ylim = c(700000,850000))
lines(fc_x_chart$`Point Forecast`, col='red')
lines(fc_x_chart_hwadd$`Point Forecast`, col='blue')
lines(fc_x_chart_hwmult$`Point Forecast`, col='green')
lines(fc_stl1_chart$`Point Forecast`, col='black')
lines(forecast_tail_chart$yhat, col='purple')
lines(e_pred_chart, col='orange')

```


#### Deep learning model using MPL - Please note that due to the time it takes to run this, we've commented the code out for purposes of printing this out into html.  Please
#### refer to slides for output of the model.
```{r}
########### Pattern Matching ############
#Deep Learning
# start.time <- Sys.time()
# 
# fit_dl <-mlp(train_ts[,"ADELAIDE.ALICE.SPRINGS"], hd.auto.type="cv")
# plot(fit_dl)
# frc2 <-forecast(fit_dl, h=12)
# print(frc2)
# plot(frc2, main = "ADELAIDE.ALICE.SPRINGS")
# 
# end.time <- Sys.time()
# time.taken <- round(end.time - start.time,2)
# time.taken
# 
# 
# #frc2
# frc2_df <- as.data.frame(frc2)
# frc2_df <- as.numeric(unlist(frc2_df))
# #frc2_df
# 
# plot(test[,"ADELAIDE.ALICE.SPRINGS"], main="ADELAIDE.ALICE.SPRINGS", xlab = "Month", ylab = "Passenger Count")  
# lines(frc2_df, col='red')
# 
# 
# #### Compare smapes of NN vs StL:
# STL_smape <- model_aa_summary %>% filter(Route == "ADELAIDE.ALICE.SPRINGS") %>% filter(Model == "STL")
# STL_smape <- STL_smape$sMAPE
# STL_smape
# 
# NN_smape <- smape(frc2_df,test[,"ADELAIDE.ALICE.SPRINGS"])
# NN_smape
# 
# #### Simple bar chart of 2 smape values:
# smape_contest <- data.frame(route = character(), model_name = character(), sMAPE = double())
# smape_contest <- rbind(smape_contest, c("ADELAIDE.ALICE.SPRINGS","STL",STL_smape))
# smape_contest <- rbind(smape_contest, c("ADELAIDE.ALICE.SPRINGS","MLP",NN_smape))
# colnames(smape_contest) <- c("Route","Model","sMAPE")
# smape_contest$sMAPE <- as.numeric(smape_contest$sMAPE)
# 
# 
# smape_contest$sMAPE <- round(smape_contest$sMAPE,4)
# 
# par(mar=c(8,6,6,6))
# yy_smape <- barplot(smape_contest$sMAPE, main="sMAPE: ADELAIDE.ALICE.SPRINGS", xlab="Model", names.arg = smape_contest$Model, las = 2, ylim = c(0,.20))
# text(x = yy_smape, y = smape_contest$sMAPE, label = smape_contest$sMAPE, pos = 3, cex = 0.8, col = "red")

```

```{r}

```

```{r}

```